---
title: "Chapter 2"
output: html_notebook
---
#Chapter 2 R basics

 In this book, we will be using the R software environment for all our analysis. You will learn R and data analysis techniques simultaneously. To follow along you will therefore need access to R. We also recommend the use of an integrated development environment(IDE), such as Rstudio, to save your work. Note that it is common for a course or workshop to offer access to an R environment and an IDE through your web browser, as done by Rstudio cloud. If you have access to such a resource, you don't need to install R and R studio. However, if you intend on becoming an advanced data analyst, we highly recommend installing these tools on your computer. Both R and Rstudio are free and available online.

##2.1 Case study: US Gun Murders

 Imagine you live in Europe and are offered a job in a US company with many locations across all states. It is a great job, but news with headines such as US Gun Homicide Rate Higher Than Other Developed Countries have you worried. Charts like this may concern you even more:
 
 Or even worse, this version from everytown.org:
 
 But then you remember that the US is a large and diverse country with 50 very different states as well as the District of Columbia (DC).
 
 California, for example, has a larger population than Canada, and 20 US states have populations larger than that of Norway. In some respects, the variability across states in the US is akin to the variability across countries in Europe. Furthermore, although not included in the charts above, the murder rates in Lithuania, Ukraine, and Russia are higher than 4 per 100,000. So perhaps the news reports that worried you are too superficial. You have options of where to live and want to determine the safety of each particular state. We will gain some insights by examining data related to gun homicides in the US during 2010 using R.

Before we get started with our example, we need to cover logistics as well as some of the very basic building blocks that are required to gain more advanced R skills. Be aware that the usefulness of some of these building blocks may not be immediately obvious, but later in the book you will appreciate having mastered these skills.

## 2.2 The very basics

 Before we get started with the motivating dataset, we need to cover the very basics of R.
 
### 2.2.1 Objects

 Suppose a high school student asks us for help solving several quadratic equations of the form 'ax^2+bx+c=0'. The quadratic formula gives us the solutions:
 
 '(-b-(b^2-4ac)^(1/2))/2a' and '(-b+(b^2-4ac)^(1/2))/2a'
 
 Which of course change depending on the values of a,b, and c. One advantage of programming languages is that we can define variables and write expressions with these variables, similar to how we do so in math, but obtain a numeric solutions. We will write out general code for the quadratic equation below, but if we are asked to solve 'x^2+x-1=0', then we define.:
 
```{r}
a <- 1
b <- 1
c <- -1

```

 which stores the values for later use. We use <- to assign values to the variables.
 
 We can also assign values using = instead of <-, but we recommend against using = to avoid confusion.
 
 Copy and paste the code above into your console to define the three variables. Note that R does not print anything when we make this assignment. This means the objects were defined successfully. Had you made a mistake, you would have received an error message.
 
 To  see the value stored in a variable, we simply ask R to evaluate 'a' and it shows the stored value:
 
```{r}
a
```
 
 A more explicit way to ask R to show us the value stored in 'a' is using 'print' like this:
 
```{r}
print(a)
```
 
 We use the term object to describe stuff that is stored in R. Variables are examples, but objects can also be more complicated entities such as functions, which are described later.
 
### 2.2.2 The workspace 

 As we define objects in the console, we are actually changing the workspace. You can see all the variables saved in your workspace by typing:
```{r}
ls()
```
 
 In RStudio, the Environment tab shows the values:
 We should see a, b, and c. If you try to recover the value of a variable that is not in your workspace, you receive an error. For example, if you type x you will receive the following message: Error: object 'x' not found.

 Now since these values are saved in variables, to obtain a solution to our equation, we use the quadratic formula:
```{r}
(-b+sqrt(b^2 -4*a*c))/(2*a)
(-b-sqrt(b^2-4*a*c))/(2*a)
```
 
### 2.2.3 Functions
 Once you define variables, the data analysis process can usually be described as a series of functions applied to the data. R includes several predefined functions and most of the analysis pipelines we construct make extensive use of these.

We already used the install.packages, library, and ls functions. We also used the function sqrt to solve the quadratic equation above. There are many more prebuilt functions and even more can be added through packages. These functions do not appear in the workspace because you did not define them, but they are available for immediate use.

In general, we need to use parentheses to evaluate a function. If you type ls, the function is not evaluated and instead R shows you the code that defines the function. If you type ls() the function is evaluated and, as seen above, we see objects in the workspace.

Unlike ls, most functions require one or more arguments. Below is an example of how we assign an object to the argument of the function log. Remember that we earlier defined a to be 1:
```{r}
log(8)
log(a)
```
You can find out what the function expects and what it does by reviewing the very useful manuals included in R. You can get help by using the help function like this:
```{r}
help("log")
```

For most functions, we can also use this shorthand:
```{r}
?log
```

The help page will show you what arguments the function is expecting. For example, log needs x and base to run. However, some arguments are required and others are optional. You can determine which arguments are optional by noting in the help document that a default value is assigned with =. Defining these is optional. For example, the base of the function log defaults to base = exp(1) making log the natural log by default.

If you want a quick look at the arguments without opening the help system, you can type:
```{r}
args(log)
```

 You can change the default values by simply assigning another object:
```{r}
log(8, base=2)
```
 Note that we have not been specifying the argument x as such:
```{r}
log(x=8, base=2)
```
 
 The above code works, but we can save ourselves some typing: if no argument name is used, R assumes you are entering arguments in the order shown in the help file or by args. So by not using the names, it assumes the arguments are x followed by base:
```{r}
log(8,2)
```
 If using the arguments’ names, then we can include them in whatever order we want:
```{r}
log(base=2, x=8)
```
 
 To specify arguments, we must use =, and cannot use <-.

There are some exceptions to the rule that functions need the parentheses to be evaluated. Among these, the most commonly used are the arithmetic and relational operators. For example:
```{r}
2^3
```
You can see the arithmetic operators by typing:
```{r}
help("+")
```
or
```{r}
?"+"
```
and the relational operators by typing:
```{r}
help(">")
```

or
```{r}
?">"
```

### 2.2.4 Other prebuilt objects
There are several datasets that are included for users to practice and test out functions. You can see all the available datasets by typing:
